---
title: "case1_FINAL"
author: "Ian Hua, InHee Ho, Sonia Xu"
date: "September 18, 2017"
output: html_document
---

```{r include = F}
library(dplyr)
library(lme4)
library(ggplot2)
library(magrittr)
#read in the data
###plot multiple ggplots in one place  obtained from http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

RMSE <- function(x,y) {return(sqrt(mean((y-x)^2)))} 
dat <- read.table("case1.txt", header = T, stringsAsFactors = F, na.strings = ".", colClasses=c("character","character","numeric","numeric","numeric","numeric","numeric","numeric"))
dat <-  dat %>% filter(complete.cases(blot)) %>% select(-wet) 
```

#Introduction
The following data is from an international validation study (19 labs, 6 nations) of the rat uterotrophic bioassay. The purpose of the assay was to screen chemicals for estrogenic effects--agonist (EE) or antagonist (ZM). One clear challenge from the assay was the consistency among variables and protocols throughout all data in the study. The motivation of this study is to fit the best model against the data to identify potential differences between labs, and to assess the consistency and effectiveness of the two chemicals. 
```{r}
```

#Methodology
Our approach to answer our problem followed these 4 steps: \newline
1. Data-cleaning \newline
2. Exploratory Data Analysis \newline
3. Fitting the data to a Linear Mixed-Effect Model \newline
4. Interpreting the data, model, and results \newline

#Dataset
The final dataset contains 2677 observations and 6 features: lab, protocol, dose 1, dose 2, body weight, and blotted uterus weight.
Three key pieces of data were removed: 4 observations due to missing data, the feature, group, due to multicollinearity, and the feature, wet, due to its variability in capturing the uterus weight. Below is a preview of the final dataset:
```{r}
head(dat)
```

##Exploratory Data Analysis
After thorough exploratory data analysis, 
```{r}
#maybe implies that we should log(blot)? idk
ggplot(data = dat, aes(y = lab, x = blot, colour = lab)) + geom_point(alpha=0.05, size=10) + theme(legend.position="none")
```
Grouping the data by lab, let us take a closer look at the relationships between dose1, dose2, body, and blot. We can easily see the points at dose1 = 10.0 results in high leverage.
```{r}
mean_dat1 <- dat %>% group_by(lab, dose1) %>% summarise(blot = mean(blot))
ggplot(data = mean_dat1, aes(x = dose1, y = blot, color = lab)) + geom_line() + ggtitle("Dose 1")
mean_dat2 <- dat %>% group_by(lab, dose2) %>% summarise(blot = mean(blot))
ggplot(data = mean_dat2, aes(x = dose2, y = blot, color = lab)) + geom_line() + ggtitle("Dose 2")
mean_body <- dat %>% group_by(lab, body) %>% summarise(blot = mean(blot))
ggplot(data = mean_body, aes(x = body, y = blot, color = lab)) + geom_point(alpha = 0.5) + ggtitle("Body") + stat_smooth(alpha = 0.1)
```
It is critical to counter the leverage in our regression model - we apply a log on these explanatory variables to account for this leverage. After, we notice that the plots are not exactly linear and it is wiser to include a polynomial to better fit the model.
```{r}
ggplot(data = dat, aes(x=(log(dose1 +1)^2), y=blot)) + geom_point()
ggplot(data = dat, aes((log(dose1 +1)^2))) + geom_histogram(fill = "blue", alpha = 0.2, binwidth = 1, col="blue") 
ggplot(data = dat, aes((log(dose2 +1)^2))) + geom_histogram(fill = "blue", alpha = 0.2, binwidth = 0.2, col="blue") 
```
Thus we use $log(dose1)^2$, $log(dose2)^2$, $log(body)^2$ in our model to fix this.
Additionally, it is likely that the protocols at each lab differ due to a multitude of reasons, including environmental, procedural, governmental, etc. We treat this as a fixed effect and include this interaction influence in our final model.


##Final Model: Linear Mixed Effect Model
The final model used to capture the variability in the effects of the two doses on uterus weight across labs and other differences in protocol between labs wa a Linear Mixed Effect Model. The random effects are dose 1 and dose 2 for each lab, and the fixed effects are dose 1, dose 2, body, and an interaction between protocol and lab. The response is the blotted uterus weight.

We noticed that adding dose1 and dose2 as both fixed and random effects better fits the data. The fixed effect captures the population effects of dose1 and dose2 across all labs and random effect captures the variation in the effects of dose1 and dose2 between labs. Similarly, we also assumed that different types of protocol have a different effect for each labs. After fitting the model with the interaction between protocol and lab (proto*lab) and with the random effects of protocol on each labs (proto|lab) respectively, we observed that the former better fits the actual dataset (see Appendix). 

To better fit the model, transformations of the features and response were utilized. Originally the response varaible was simply blot - this was incorrect due to differences in scale in addition to heteroskedasticity in the model. This was corrected in the final model using $log(blot+1)$ as the response variable, which ______.
```{r}
fm <- lmer(log(blot+1) ~ (1 + poly(log(dose1 + 1),2) + poly(log(dose2 + 1),2) |lab) + poly(log(dose1 + 1),2) + poly(log(dose2+1),2) + poly(log(body),2) + (proto*lab), dat)
plot(fm, main = "Fitted vs. Residuals", xlab = "fitted", ylab = "residuals")
```


#Discussion of Final Model
```{r}
```
-normality
-residual diagnostics
-every linear check possible
-what can we say about the variability between labs?

#Recommendations
-scientific implications of the model
```{r}
```

#Appendix
